---
title: "EDA"
author: "Andy Powers"
format: html
---

# Project 3: Modeling

This project explores a dataset, develops models for it, and provides an API interface for on-demand model results. This is one of two Quarto documents; the other is linked at the bottom of this page.

## Project introduction

### Dataset

We are studying data on diabetes health indicators, pulled from Kaggle. The following summaries are from the [dataset source page](https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/):

#### Survey summary

*The Behavioral Risk Factor Surveillance System (BRFSS) is a health-related telephone survey that is collected annually by the CDC. Each year, the survey collects responses from over 400,000 Americans on health-related risk behaviors, chronic health conditions, and the use of preventative services. It has been conducted every year since 1984. For this project, a csv of the dataset available on Kaggle for the year 2015 was used. This original dataset contains responses from 441,455 individuals and has 330 features. These features are either questions directly asked of participants, or calculated variables based on individual participant responses.*

#### Data file summary

`diabetes _ binary _ health _ indicators _ BRFSS2015.csv` *is a clean dataset of 253,680 survey responses to the CDC's BRFSS2015. The target variable Diabetes_binary has 2 classes. 0 is for no diabetes, and 1 is for prediabetes or diabetes. This dataset has 21 feature variables and is not balanced.*

#### Column descriptions

`Diabetes_binary` **(response variable)**
0 = no diabetes 1 = diabetes

`HighBP`
0 = no high BP 1 = high BP

`HighChol`
0 = no high cholesterol 1 = high cholesterol

`CholCheck`
0 = no cholesterol check in 5 years 1 = yes cholesterol check in 5 years

`BMI`
Body Mass Index

`Smoker`
Have you smoked at least 100 cigarettes in your entire life? [Note: 5 packs = 100 cigarettes] 0 = no 1 = yes

`Stroke`
(Ever told) you had a stroke. 0 = no 1 = yes

`HeartDiseaseorAttack`
coronary heart disease (CHD) or myocardial infarction (MI) 0 = no 1 = yes

`PhysActivity`
physical activity in past 30 days - not including job 0 = no 1 = yes

`Fruits`
Consume Fruit 1 or more times per day 0 = no 1 = yes

`Veggies`
Consume Vegetables 1 or more times per day 0 = no 1 = yes

`HvyAlcoholConsump`
(adult men >=14 drinks per week and adult women>=7 drinks per week) 0 = no 1 = yes

`AnyHealthcare`
Have any kind of health care coverage, including health insurance, prepaid plans such as HMO, etc. 0 = no 1 = yes

`NoDocbcCost`
Was there a time in the past 12 months when you needed to see a doctor but could not because of cost? 0 = no 1 = yes

`GenHlth`
Would you say that in general your health is: scale 1-5 1 = excellent 2 = very good 3 = good 4 = fair 5 = poor

`MentHlth`
days of poor mental health scale 1-30 days

`PhysHlth`
physical illness or injury days in past 30 days scale 1-30

`DiffWalk`
Do you have serious difficulty walking or climbing stairs? 0 = no 1 = yes

`Sex`
0 = female 1 = male

`Age`
13-level age category (_AGEG5YR see codebook) scale 1-13 
- 1 = 18-24 
- 2 = 25-29 
- 3 = 30-34 
- 4 = 35-39 
- 5 = 40-44 
- 6 = 45-49 
- 7 = 50-54 
- 8 = 55-59 
- 9 = 60-64 
- 10 = 65-69 
- 11 = 70-74 
- 12 = 75-79 
- 13 = 80 or older 

`Education`
Education level (EDUCA see codebook) scale 1-6 
- 1 = Never attended school or only kindergarten 
- 2 = Grades 1 through 8 (Elementary) 
- 3 = Grades 9 through 11 (Some high school) 
- 4 = Grade 12 or GED (High school graduate) 
- 5 = College 1 year to 3 years (Some college or technical school) 
- 6 = College 4 years or more (College graduate)

`Income`
Income scale (INCOME2 see codebook) scale 1-8 
- 1 = less than $10,000 
- 2 = Less than $15,000 ($10,000 to less than $15,000) 
- 3 = Less than $20,000 ($15,000 to less than $20,000) 
- 4 = Less than $25,000 ($20,000 to less than $25,000) 
- 5 = Less than $35,000 ($25,000 to less than $35,000) 
- 6 = Less than $50,000 ($35,000 to less than $50,000) 
- 7 = Less than $75,000 ($50,000 to less than $75,000) 
- 8 = $75,000 or more

### Contents

There are six components to this project.

1. EDA - quarto file with EDA *(this file)*

1. Modeling - quarto file to select the best model using `tidymodels`

1. API - R file fitting the model and defining API structure

1. Dockerfile - for building the Docker image

1. .tar - Docker image

1. Demo - video showing me demonstrating my Docker container

### R libraries

```{r}
library(tidyverse)
library(tidymodels)

```

## EDA

As typical, let's conduct standard data cleanup checks, ETL, and other EDA tasks before proceeding to model the data.

- proper data types
- missing or malformed values
- factoring
- categorical summaries (single and multi)
- numeric summaries (single and multi)
- plots

### Data read

Read the data into our instance, in the variable `data`.

```{r}
raw_data <- read_csv(file = 'diabetes_binary_health_indicators_BRFSS2015.csv')
data <- raw_data #saved for backup if needed
```

### Clean any NA

Check for and handle any NA found.

```{r}
sum_na <- function (column) { sum(is.na(column)) }
data |> summarize(across(everything(), sum_na))
```

None found. Moving on.

### Clean up data types

Review the data types auto-assigned to each column. We know that there will be categorical variables identified as numerics, requiring factor updates.

```{r}
str(data)
summary(data)
```

We need to factor many of these fields. I also like to change doubles to integers for appropriate numeric fields. I checked the summary() to confirm that categorical variables like `Age` do not have values outside of the known ranges.

#### Converting factors

First, factoring the simple ones with only a 0/1, then the batch of larger categorical fields.

```{r}
#simple factors first
data <- data |> mutate(
  Diabetes_binary = factor(Diabetes_binary,labels=c("not diabetic","diabetic")),
  HighBP = factor(HighBP,labels=c("no high blood pressure","high blood pressure")),
  HighChol = factor(HighChol,labels=c("no high cholesterol","high cholesterol")),
  CholCheck = factor(CholCheck,labels=c("no recent cholesterol check","recent cholesterol check")),
  Smoker = factor(Smoker,labels=c("non-smoker","smoker")),
  Stroke = factor(Stroke,labels=c("no stroke","stroke")),
  HeartDiseaseorAttack = factor(HeartDiseaseorAttack,labels=c("no heart problems CHD/MI","heart problems CHD/MI")),
  PhysActivity = factor(PhysActivity,labels=c("no physical activity","physical activity")),
  Fruits = factor(Fruits,labels=c("no fruits","fruits")),
  Veggies = factor(Veggies,labels=c("no vegetables","vegetables")),
  HvyAlcoholConsump = factor(HvyAlcoholConsump,labels=c("no heavy alcohol","heavy alcohol")),
  AnyHealthcare = factor(AnyHealthcare,labels=c("no healthcare","healthcare")),
  NoDocbcCost = factor(NoDocbcCost,labels=c("not avoided doctor for cost","avoided doctor for cost")),
  DiffWalk = factor(DiffWalk,labels=c("no walking difficulty","walking difficulty")),
  Sex = factor(Sex,labels=c("female","male"))
)

#larger categories
data <- data |> mutate(
  GenHlth = factor(GenHlth,labels=c("excellent","very good","good","fair","poor")),
  Age = factor(Age,labels=c(
    "18-24",
    "25-29",
    "30-34",
    "35-39",
    "40-44",
    "45-49",
    "50-54",
    "55-59",
    "60-64",
    "65-69",
    "70-74",
    "75-79",
    "80+")),
  Education = factor(Education,labels=c(
    "Never attended school or only kindergarten",
    "Grades 1-8 (Elementary)",
    "Grades 9-11 (Some high school)",
    "Grades 12 or GED (High school graduate)",
    "College 1-3 years (Some college or technical school",
    "College 4+ years (College graduate)")),
  Income = factor(Income,labels=c(
    "[ - $10k)",
    "[$10k - $15k)",
    "[$15k - $20k)",
    "[$20k - $25k)",
    "[$25k - $35k)",
    "[$35k - $50k)",
    "[$50k - $75k)",
    "[$75k - ]"))
)
  
```



`GenHlth`
Would you say that in general your health is: scale 1-5 1 = excellent 2 = very good 3 = good 4 = fair 5 = poor

`DiffWalk`
Do you have serious difficulty walking or climbing stairs? 0 = no 1 = yes

`Sex`
0 = female 1 = male

`Age`
13-level age category (_AGEG5YR see codebook) scale 1-13 
- 1 = 18-24 
- 2 = 25-29 
- 3 = 30-34 
- 4 = 35-39 
- 5 = 40-44 
- 6 = 45-49 
- 7 = 50-54 
- 8 = 55-59 
- 9 = 60-64 
- 10 = 65-69 
- 11 = 70-74 
- 12 = 75-79 
- 13 = 80 or older 

`Education`
Education level (EDUCA see codebook) scale 1-6 
- 1 = Never attended school or only kindergarten 
- 2 = Grades 1 through 8 (Elementary) 
- 3 = Grades 9 through 11 (Some high school) 
- 4 = Grade 12 or GED (High school graduate) 
- 5 = College 1 year to 3 years (Some college or technical school) 
- 6 = College 4 years or more (College graduate)

`Income`
Income scale (INCOME2 see codebook) scale 1-8 
- 1 = less than $10,000 
- 2 = Less than $15,000 ($10,000 to less than $15,000) 
- 3 = Less than $20,000 ($15,000 to less than $20,000) 
- 4 = Less than $25,000 ($20,000 to less than $25,000) 
- 5 = Less than $35,000 ($25,000 to less than $35,000) 
- 6 = Less than $50,000 ($35,000 to less than $50,000) 
- 7 = Less than $75,000 ($50,000 to less than $75,000) 
- 8 = $75,000 or more




[Click here for the Modeling Page](Modeling.html)
