---
title: "EDA"
author: "Andy Powers"
format: html
---

# Project 3: Modeling

This project explores a dataset, develops models for it, and provides an API interface for on-demand model results. This is one of two Quarto documents; the other is linked at the bottom of this page.

## Project introduction

### Dataset

We are studying data on diabetes health indicators, pulled from Kaggle. The following summaries are from the [dataset source page](https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/):

#### Survey summary

*The Behavioral Risk Factor Surveillance System (BRFSS) is a health-related telephone survey that is collected annually by the CDC. Each year, the survey collects responses from over 400,000 Americans on health-related risk behaviors, chronic health conditions, and the use of preventative services. It has been conducted every year since 1984. For this project, a csv of the dataset available on Kaggle for the year 2015 was used. This original dataset contains responses from 441,455 individuals and has 330 features. These features are either questions directly asked of participants, or calculated variables based on individual participant responses.*

#### Data file summary

`diabetes _ binary _ health _ indicators _ BRFSS2015.csv` *is a clean dataset of 253,680 survey responses to the CDC's BRFSS2015. The target variable Diabetes_binary has 2 classes. 0 is for no diabetes, and 1 is for prediabetes or diabetes. This dataset has 21 feature variables and is not balanced.*

#### Column descriptions

`Diabetes_binary` **(response variable)** 0 = no diabetes 1 = diabetes

`HighBP` 0 = no high BP 1 = high BP

`HighChol` 0 = no high cholesterol 1 = high cholesterol

`CholCheck` 0 = no cholesterol check in 5 years 1 = yes cholesterol check in 5 years

`BMI` Body Mass Index

`Smoker` Have you smoked at least 100 cigarettes in your entire life? \[Note: 5 packs = 100 cigarettes\] 0 = no 1 = yes

`Stroke` (Ever told) you had a stroke. 0 = no 1 = yes

`HeartDiseaseorAttack` coronary heart disease (CHD) or myocardial infarction (MI) 0 = no 1 = yes

`PhysActivity` physical activity in past 30 days - not including job 0 = no 1 = yes

`Fruits` Consume Fruit 1 or more times per day 0 = no 1 = yes

`Veggies` Consume Vegetables 1 or more times per day 0 = no 1 = yes

`HvyAlcoholConsump` (adult men \>=14 drinks per week and adult women\>=7 drinks per week) 0 = no 1 = yes

`AnyHealthcare` Have any kind of health care coverage, including health insurance, prepaid plans such as HMO, etc. 0 = no 1 = yes

`NoDocbcCost` Was there a time in the past 12 months when you needed to see a doctor but could not because of cost? 0 = no 1 = yes

`GenHlth` Would you say that in general your health is: scale 1-5 1 = excellent 2 = very good 3 = good 4 = fair 5 = poor

`MentHlth` days of poor mental health scale 1-30 days

`PhysHlth` physical illness or injury days in past 30 days scale 1-30

`DiffWalk` Do you have serious difficulty walking or climbing stairs? 0 = no 1 = yes

`Sex` 0 = female 1 = male

`Age` 13-level age category (\_AGEG5YR see codebook) scale 1-13 - 1 = 18-24 - 2 = 25-29 - 3 = 30-34 - 4 = 35-39 - 5 = 40-44 - 6 = 45-49 - 7 = 50-54 - 8 = 55-59 - 9 = 60-64 - 10 = 65-69 - 11 = 70-74 - 12 = 75-79 - 13 = 80 or older

`Education` Education level (EDUCA see codebook) scale 1-6 - 1 = Never attended school or only kindergarten - 2 = Grades 1 through 8 (Elementary) - 3 = Grades 9 through 11 (Some high school) - 4 = Grade 12 or GED (High school graduate) - 5 = College 1 year to 3 years (Some college or technical school) - 6 = College 4 years or more (College graduate)

`Income` Income scale (INCOME2 see codebook) scale 1-8 - 1 = less than \$10,000 - 2 = Less than \$15,000 (\$10,000 to less than \$15,000) - 3 = Less than \$20,000 (\$15,000 to less than \$20,000) - 4 = Less than \$25,000 (\$20,000 to less than \$25,000) - 5 = Less than \$35,000 (\$25,000 to less than \$35,000) - 6 = Less than \$50,000 (\$35,000 to less than \$50,000) - 7 = Less than \$75,000 (\$50,000 to less than \$75,000) - 8 = \$75,000 or more

### Contents

There are six components to this project.

1.  EDA - quarto file with EDA *(this file)*

2.  Modeling - quarto file to select the best model using `tidymodels`

3.  API - R file fitting the model and defining API structure

4.  Dockerfile - for building the Docker image

5.  .tar - Docker image

6.  Demo - video showing me demonstrating my Docker container

### R libraries

```{r}
library(tidyverse)
library(tidymodels)

```

## EDA

As typical, let's conduct standard data cleanup checks, ETL, and other EDA tasks before proceeding to model the data. Although a best practice would be to split the data at hand into a training and testing set first, I'm studying the full data.

-   missing or malformed values
-   proper data types
-   factoring
-   categorical summaries
-   numeric summaries

### Data read

Read the data into our instance, in the variable `data`.

```{r}
raw_data <- read_csv(file = 'diabetes_binary_health_indicators_BRFSS2015.csv')
data <- raw_data #saved for backup if needed
```

### Clean any NA

Check for and handle any NA found.

```{r}
sum_na <- function (column) { sum(is.na(column)) }
data |> summarize(across(everything(), sum_na))
```

None found. Moving on.

### Clean up data types

Review the data types auto-assigned to each column. We know that there will be categorical variables identified as numerics, requiring factor updates.

```{r}
str(data)
summary(data)
```

We need to factor many of these fields. I also like to change doubles to integers for appropriate numeric fields. I checked the summary() to confirm that categorical variables like `Age` do not have values outside of the known ranges.

#### Converting factors

First, factoring the simple ones with only a 0/1, then the batch of larger categorical fields.

```{r}
#simple factors first
data <- data |> mutate(
  Diabetes_binary = factor(Diabetes_binary,labels=c("not diabetic","diabetic")),
  HighBP = factor(HighBP,labels=c("no high blood pressure","high blood pressure")),
  HighChol = factor(HighChol,labels=c("no high cholesterol","high cholesterol")),
  CholCheck = factor(CholCheck,labels=c("no recent cholesterol check","recent cholesterol check")),
  Smoker = factor(Smoker,labels=c("non-smoker","smoker")),
  Stroke = factor(Stroke,labels=c("no stroke","stroke")),
  HeartDiseaseorAttack = factor(HeartDiseaseorAttack,labels=c("no heart problems CHD/MI","heart problems CHD/MI")),
  PhysActivity = factor(PhysActivity,labels=c("no physical activity","physical activity")),
  Fruits = factor(Fruits,labels=c("no fruits","fruits")),
  Veggies = factor(Veggies,labels=c("no vegetables","vegetables")),
  HvyAlcoholConsump = factor(HvyAlcoholConsump,labels=c("no heavy alcohol","heavy alcohol")),
  AnyHealthcare = factor(AnyHealthcare,labels=c("no healthcare","healthcare")),
  NoDocbcCost = factor(NoDocbcCost,labels=c("not avoided doctor for cost","avoided doctor for cost")),
  DiffWalk = factor(DiffWalk,labels=c("no walking difficulty","walking difficulty")),
  Sex = factor(Sex,labels=c("female","male"))
)

#larger categories
data <- data |> mutate(
  GenHlth = factor(GenHlth,labels=c("excellent","very good","good","fair","poor")),
  Age = factor(Age,labels=c(
    "18-24",
    "25-29",
    "30-34",
    "35-39",
    "40-44",
    "45-49",
    "50-54",
    "55-59",
    "60-64",
    "65-69",
    "70-74",
    "75-79",
    "80+")),
  Education = factor(Education,labels=c(
    "Never attended school or only kindergarten",
    "Grades 1-8 (Elementary)",
    "Grades 9-11 (Some high school)",
    "Grades 12 or GED (High school graduate)",
    "College 1-3 years (Some college or technical school",
    "College 4+ years (College graduate)")),
  Income = factor(Income,labels=c(
    "[ - $10k)",
    "[$10k - $15k)",
    "[$15k - $20k)",
    "[$20k - $25k)",
    "[$25k - $35k)",
    "[$35k - $50k)",
    "[$50k - $75k)",
    "[$75k - ]"))
)
  
```

#### Adjust numeric

Transitioning from double to integers where needed. Checking values to confirm integers appropriate before switch.

```{r}
#checking results for integer reasonableness
unique(data$BMI)
unique(data$MentHlth)
unique(data$PhysHlth)

#convert doubles to ints
data <- data |> mutate(
  BMI = as.integer(BMI),
  MentHlth = as.integer(MentHlth),
  PhysHlth = as.integer(PhysHlth)
)

str(data)
```

### Explore summaries

Making a few exploratory charts and plots, to increase understanding of the data and explore hunches and questions.

#### Categorical

This section explores a few categorical variables with contingency tables and bar plots.

What is our distribution of failures and successes (Diabetes_binary)?

```{r}
data |> 
  group_by(Diabetes_binary) |>
  summarize(count = n())
```

About 7:1 non-diabetic vs diabetic, with satisfactory sample sizes.

Let's explore the sex and age makeups. I'm switching to plots because I find the tabular, numeric contingency data to be less informative for exploration.

```{r}
g <- data |>
  ggplot(
    aes(
      x=Age
    )
  ) +
  labs(
    title="Age Group Representation"
  ) +
  geom_bar(
    
  )
g

```

Most in their 50-60s, generally older subjects.

```{r}
g <- data |>
  ggplot(
    aes(
      x=Age,
      fill=Sex
    )
  ) +
  labs(
    title="Age Group Representation by Sex"
  ) +
  geom_bar(

  )
g

```

Reasonably equal makeup of men and women across age groups. How about the same faceted across income levels?

```{r}
g <- data |>
  ggplot(
    aes(
      x=Age,
      fill=Sex
    )
  ) +
  labs(
    title="Age Group Representation by Sex faceted by Income",
    x=''
  ) +
  geom_bar(

  ) +
  facet_grid(
    . ~ Income
  ) +
  theme(
    axis.text.x=element_blank()
  )
g

```

Our survey data leans heavily on middle class+ respondents. It's worth recalling this as interpretations are made in aggregate, either to temper them with a known skew toward high earners or else grouped by income levels as a known strata of interest. Then again, this variable may have no significant bearing on the predictive model, in which case this point is unnecessary.

Let's adjust this to see diabetic state by income level.

```{r}
g <- data |>
  ggplot(
    aes(
      x=Age,
      fill=Diabetes_binary
    )
  ) +
  labs(
    title="Age Group Representation by Diabetes state faceted by Income",
    x=''
  ) +
  geom_bar(

  ) +
  facet_grid(
    . ~ Income
  ) +
  theme(
    axis.text.x=element_blank()
  )
g

```

At a glance, I think the proportion of diabetic respondents is higher for lower income levels. Interesting but no takeaway right now.

#### Numerical

We have three numeric fields to explore: BMI, Mental Health, and Physical Health. The encoding here is a little confusing, where higher numbers indicate more detrimental values of BMI, mental, or physical health. The two health fields are a count of problems experienced in the last X days.

First, let's check density charts for Diabetic and non-Diabetic fills.

```{r}
g <- data |>
  ggplot(
    aes(
      x=BMI,
      fill=Diabetes_binary
    )
  ) +
  labs(
    title="BMI density chart with Diabetes state",
  ) +
  geom_density(
    alpha=0.5
    #bindwidth=11
  ) 
g
```

That is staggeringly clear, a strong predictor.

```{r}
g <- data |> ggplot()
g +
  aes(
    x=MentHlth,
    fill=Diabetes_binary
  ) +
  labs(
    title="Mental Health Incidents Histogram with Diabetes state",
  ) +
  geom_histogram(
    alpha=0.5,
    binwidth=2
  ) 

```

```{r}
g <- data |> ggplot()
g +
  aes(
    x=PhysHlth,
    fill=Diabetes_binary
  ) +
  labs(
    title="Physical Health Incidents Histogram with Diabetes state",
  ) +
  geom_histogram(
    alpha=0.5,
    binwidth=2
  ) 

```

These health numeric variables are skewed heavily right, with most respondents reporting 0. I don't expect much to come from it but let's check summaries and correlation too.

```{r}
data |> 
  select(BMI,MentHlth,PhysHlth) |>
  summary()
```

```{r}
cor(data[sapply(data,is.numeric)]) 
```

The summary stats confirm the takes noted above. Stronger correlation between mental health incident days and physical health incident days, which sounds reasonable. No need to explore a scatterplot here, for these few and clear numeric variables.

Instead, let's check a boxplot of the numeric fields across categories including the response variable.

```{r}
g <- data |> ggplot()
g +
  aes(

  ) +
  labs(
    title="BMI over Diabetes state by Sex",
  ) +
  geom_boxplot(
    aes(
      x=BMI,
      fill=Diabetes_binary
      )  
    ) +
  facet_grid( Sex~.)
```

So, higher BMIs link with diabetic state in both sexes. Also, interesting to see the range of BMI for diabetic females is a little wider than that of males.

Done with EDA here. I think many of the predictors may relate to the response variable, so I'll proceed to modeling to understand more about the stronger predictor variables, then perhaps study further. I'll save and load the updated tibble with my data for processing in the Modeling page.

[Click here for the Modeling Page](Modeling.html)

```{r}
#here, just saving data frame for use in Modeling
save(data,file="data.RData")
```

